Neutron Star Mergers and their Impact on Second Generation Star Formation in the Early Universe

### 2023-09-14 NSM_Skinner_Wise/NSM_mixing.tex
Minor changes to new text

% Abstract of the paper
\begin{abstract}
	The exact evolution of elements in the universe, from primordial hydrogen and helium to heavier elements like gold and platinum produced via the r-process, is still under scrutiny. The supernova deaths of the very first stars to form in the universe led to the enrichment of their local environments with new metals, and can leave behind neutron stars as remnants. These remnants can end up in binary systems with other neutron stars, and eventually merge, allowing for the r-process to occur. In this work, we study the scenario where a single neutron star merger (NSM) enriches a halo early in its evolution to understand the impact on the second generation of stars and their metal abundances. We perform a suite of high resolution cosmological zoom-in simulations using \enzo{} where we have implemented a new NSM model varying the explosion energy and the delay time. In general, a NSM leads to a significant r-process enhancement in the second generation of stars. \textbf{A high explosion energy leads to a Pop II mass fraction of 72\% being highly enhanced with r-process elements, while a lower explosion energy leads to a Pop II mass fraction of 80\% being r-process enhanced, but only 14\% being highly enhanced. When the NSM has a short delay time of 10 Myr, only 5\% of the mass fraction of Pop II stars is highly r-process enhanced, while 64\% of the mass fraction of Pop II stars is highly r-process enhanced for the longest delay time of 100 Myr.} This work represents a stepping stone towards understanding how NSMs impact their environments and metal abundances of descendant generations of stars.
\end{abstract}

@@

A second explosion goes off with $E_{\textrm{NSM}}$ after a time $t_{\textrm{merge}}$, with the parameters chosen in a particular run given the choices in Table \ref{tab:NSM_params}. The final remnant particle is then converted to an inert particle with a mass equal to the sum of the remnant masses. We do not consider any EM radiation from the NSs nor any kilonova effects of a NSM for simplicity; we only follow the mass ejected from this event. We track the r-process material in a passive scalar field which does not affect metal cooling rates. For this proof of concept, the total ejecta mass is set to $0.01 \Ms$, but in practice, due to the Cartesian discretization of a sphere in of Enzo, the actual mass ejected is $0.0169 M_{\odot}$. \textbf{The delay times chosen for these runs are on the shorter end of timescales for the merging of two NSs to take place, but for this proof of concept, we wanted to explore this particular situation where a NSM occurs early in the star formation history of the galaxy. For a galaxy like Ret II, a NSM would have had to occur within 500 Myr to allow for ample time for thorough mixing to take place within the halo \citep{Simon23}. Delay times of 10, 100, and 300 Myr are short, but are not completely out of the realm of possibilities, and allow us to see how a NSM affects subsequent star formation at this early point in time.}

@@

\textbf{For all runs, we see similar Pop III stellar masses as compared to some recent studies. \citet{Riaz22} studied the Pop III contribution to high redshift galaxies for a range of halo masses. They find that a halo with a mass of $10^7$ \Ms has a Pop III mass fraction of $\sim 10^{-1}$ at $z \approx 12$. For our runs, at a similar time and halo mass, we also see a Pop III mass fraction of $\sim 10^{-1}$. By $z = 10$, we see $\sim 10^{4}$ \Ms of Pop II stars and $\sim 10^{3}$ \Ms for Pop III stars, as compared to $10^3 - 10^5$ \Ms of Pop II stars and  $10^1 - 10^2$ \Ms of Pop III stars found by \citet{Riaz22}. While we do see a slightly higher value of Pop III stellar masses as compared to them, our overall mass fractions are similar. \citet{Yajima23} studied the first galaxies and their Pop III contributions in order to compare with results from JWST. They also look at the Pop III mass fraction (see their Figure 5) as compared to the total stellar mass. They report a much broader range of values for different redshifts in general. For a stellar mass of $10^{4.5}$ \Ms, they find a Pop III mass fraction of $\sim 10^{-0.75}$, while we see a mass fraction for that stellar mass closer to $10^{-1.5}$ \Ms. However, we do see a similar trend, that as stellar mass increases, the Pop III mass fraction begins to decrease. At the end of our simulations when our halo has a stellar mass of $\sim 10^{5.5}$ \Ms, we see a Pop III mass fraction of $\sim 10^{-2.25}$, which is similar to \citet{Yajima23}. Based on these results, we agree with their conclusion that for halos with stellar masses of $\sim 10^{7}$ \Ms, the Pop III mass fraction is small, at $< 0.01$ }

This work represents some of the first simulations on the chemical and dynamical impacts of a NSM, with various parameters, in a cosmological setting, and thus there are some difficulties that come with this kind of study. The first is that the small sample size prevents us from saying anything conclusive about the NSM parameters in general. While we see some trends across the data, the inherent randomness across simulations leads to different amounts of star formation at different times and locations. On top of that, because we randomly sample the Pop III IMF, if two simulations are looking very similar and both go on to form a Pop III star in a similar location, one can be sampled at a low mass and the other at a very high mass, and affect the surrounding region in completely different ways. What may seem like a small difference in the Pop III stellar mass can go on to have impactful consequences on future star formation. This means that it can be difficult to pick apart why certain stars form where and at what times. We stress that this is merely a proof-of-concept study to show that we can study the impacts of NSMs in a cosmological setting to present meaningful results that can enhance our understanding of r-process enrichment in the early universe. \textbf{Studies are currently underway to try to separate the source of differences between the simulations, and are outside the scope of this paper.}

>>>>

% Abstract of the paper
\begin{abstract}
	The exact evolution of elements in the universe, from primordial hydrogen and helium to heavier elements like gold and platinum produced via the r-process, is still under scrutiny. The supernova deaths of the very first stars to form in the universe led to the enrichment of their local environments with new metals, and can leave behind neutron stars as remnants. These remnants can end up in binary systems with other neutron stars, and eventually merge, allowing for the r-process to occur. In this work, we study the scenario where a single neutron star merger (NSM) enriches a halo early in its evolution to understand the impact on the second generation of stars and their metal abundances. We perform a suite of high resolution cosmological zoom-in simulations using \enzo{} where we have implemented a new NSM model varying the explosion energy and the delay time. In general, a NSM leads to a significant r-process enhancement in the second generation of stars \textbf{in a galaxy with a stellar mass of $\sim 10^5 \Ms$ at redshift 10.  A high explosion energy leads to a Pop II mass fraction of 72\% being highly enhanced with r-process elements, while a lower explosion energy leads to a Pop II mass fraction of 80\% being r-process enhanced, but only 14\% being highly enhanced. When the NSM has a short delay time of 10 Myr, only 5\% of the mass fraction of Pop II stars is highly r-process enhanced, while 64\% of the mass fraction of Pop II stars is highly r-process enhanced for the longest delay time of 100 Myr.} This work represents a stepping stone towards understanding how NSMs impact their environments and metal abundances of descendant generations of stars.
\end{abstract}

@@

A second explosion goes off with $E_{\textrm{NSM}}$ after a time $t_{\textrm{merge}}$, with the parameters chosen in a particular run given the choices in Table \ref{tab:NSM_params}. The final remnant particle is then converted to an inert particle with a mass equal to the sum of the remnant masses. We do not consider any EM radiation from the NSs nor any kilonova effects of a NSM for simplicity; we only follow the mass ejected from this event. We track the r-process material in a passive scalar field which does not affect metal cooling rates. For this proof of concept, the total ejecta mass is set to $0.01 \Ms$, but in practice, due to the Cartesian discretization of a sphere in of Enzo, the actual mass ejected is $0.0169 M_{\odot}$. \textbf{The delay times chosen for these runs are on the shorter end of timescales for the merging of two NSs to take place, but for this proof of concept, we explore this particular situation where a NSM occurs early in the star formation history of the galaxy. For a galaxy like Ret II, a NSM would have had to occur within 500 Myr to allow for ample time for thorough mixing to take place within the halo \citep{Simon23}. Delay times of 10, 100, and 300 Myr are short, but are not completely out of the realm of possibilities, and allow us to see how a NSM affects subsequent star formation at this early point in time.}

@@

\textbf{For all runs, we see similar Pop III stellar masses as compared to some recent studies. \citet{Riaz22} studied the Pop III contribution to high redshift galaxies for a range of halo masses. They find that a halo with a mass of $10^7$ \Ms{} has a Pop III mass fraction of $\sim 0.1$ at $z \approx 12$. For our runs, at a similar time and halo mass, we also see a Pop III mass fraction of $\sim 0.1$. By $z = 10$, we see $\sim 10^{4}$ \Ms{} of Pop II stars and $\sim 10^{3}$ \Ms{} for Pop III stars, as compared to $10^3 - 10^5$ \Ms{} of Pop II stars and  $10^1 - 10^2$ \Ms{} of Pop III stars found by \citet{Riaz22}. While we do see a slightly higher value of Pop III stellar masses as compared to them, our overall mass fractions are similar. \citet{Yajima23} studied the first galaxies and their Pop III contributions in order to compare with results from JWST. They also look at the Pop III mass fraction (see their Figure 5) as compared to the total stellar mass. They report a much broader range of values for different redshifts in general. For a stellar mass of $10^{4.5}$ \Ms, they find a Pop III mass fraction of $\sim 0.18$, while we see a mass fraction for that stellar mass around 0.03. However, we do see a similar trend, that as stellar mass increases, the Pop III mass fraction begins to decrease. At the end of our simulations when our halo has a stellar mass of $\sim 10^{5.5}$ \Ms, we find a Pop III mass fraction of $\sim 0.006$, which is similar to \citet{Yajima23}. Based on these results, we agree with their conclusion that for halos with stellar masses of $\sim 10^{7}$ \Ms, the Pop III mass fraction is small, at $< 0.01$ }

This work represents some of the first simulations on the chemical and dynamical impacts of a NSM, with various parameters, in a cosmological setting, and thus there are some difficulties that come with this kind of study. The first is that the small sample size prevents us from saying anything conclusive about the NSM parameters in general. While we see some trends across the data, the inherent randomness across simulations leads to different amounts of star formation at different times and locations. On top of that, because we randomly sample the Pop III IMF, if two simulations are looking very similar and both go on to form a Pop III star in a similar location, one can be sampled at a low mass and the other at a very high mass, and affect the surrounding region in completely different ways. What may seem like a small difference in the Pop III stellar mass can go on to have impactful consequences on future star formation. This means that it can be difficult to pick apart why certain stars form where and at what times. We stress that this is merely a proof-of-concept study to show that we can study the impacts of NSMs in a cosmological setting to present meaningful results that can enhance our understanding of r-process enrichment in the early universe. \textbf{We plan to follow up this paper with a study that separates the source of differences between the simulations, and are outside the scope of this paper.}
